---
title: "EMS"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{EMS}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```


## Hasse Diagrams

This document demonstrates how expected mean squares can be calculated from the Hasse diagram, 
following the procedures described in [Oehlert(2010)](http://users.stat.umn.edu/~gary/book/fcdae.pdf).

```{r setup}
library(Hasse)
library(DiagrammeR)
```

## Two Way Factorial Anova

Consider the two way factorial. For concreteness, suppose A has 5 levels and B has 3, so there are
$5 \times 3 = 15$ treatment combinations in total, and suppose there are two replications of each
treatment combination so there are $5 \times 3 \times 2 = 30$ units.

```{r}
df <- expand.grid(A=factor(1:5),B=factor(1:3),Rep=factor(1:2))
```


The Hasse diagram is laid out from top to bottom, with a term for the grand mean M at the top, and a
term for the error at the bottom.  Then the terms of the model are laid in rows by order: main
effects first, then second order interactions, then third order interactions, and so on.  If a term
is nested in another, it is drawn directly under it on the next row down.

For each term, a line is drawn to every term in the row up that it "contains". A term Y is above X
if Y can be reached from X by moving up line segments, and Y is below X if Y can be reached from X
by moving down line segments.

Random factors are drawn in parentheses, and if a term is below a term with parentheses it is also
drawn in parentheses.  The error term is always random, so it is always parenthesized.


### Random A, B

If A and B are both random

```{r}
grViz(hasseDiagram(~ A*B,random=c("A","B"),data=df))
```

To calculate superscripts and subscripts

* the grand mean has both subscript and superscript 1, and the error has the number of observations
  as its superscript,
* the superscript of a term is the product of the number of levels of the factors contained in that
  term, and
* the subscript is its superscript minus the subscripts of all terms above it.

So for the diagram above

* A has $5$ levels, so it has superscript $5$ and its subscript is $5-1=4$.
* B has $3$ levels, so it has superscript $3$ and its subscript is $3-1=2$.
* A:B is constructed from A and B so its superscript is the product of the superscripts of A and B,
  $5 \times 3 = 15$, and its subscript is $15-1-4-2=8$.
* the error term always has the total number of observations as its superscript, and its subscript 
  is $30-1-4-2-8=15$.


A term X is tested against the "highest" parenthesized term below it *that does not contain a fixed
term that is not contained X* If there is more than one "highest" term, then a quasi-ratio must be
constructed.

In this case both A and B are tested against A:B, and A:B is tested against the error term.

The expected mean square for a term X is constructed from X and all the parenthesized terms below X
*that do not contain fixed terms that are not contained in X*. The coefficients are the ratios of
the superscript for the error term and the superscript for the parenthesized term.

The expected mean square for A is 
$$
\frac{30}{5} \sigma_{\alpha}^{2} + \frac{30}{15} \sigma_{\alpha\beta}^2 + \sigma^2
$$
the expected mean square for B is 
$$
\frac{30}{3} \sigma_{\beta}^{2} + \frac{30}{15} \sigma_{\alpha\beta}^2 + \sigma^2
$$
and the expected mean square for A:B is 
$$
\frac{30}{15} \sigma_{\alpha\beta}^2 + \sigma^2
$$

### Fixed A, Random B

If instead, A is fixed, all that changes are the parentheses.

```{r}
grViz(hasseDiagram(~ A*B,random=c("B"),data=df))
```

In this case A is fixed so it has no parentheses, but B is random so it is parenthesized, and AB
contains B so it is parenthesized too.

In this case, A is again tested against A:B and A:B against the error term, but A:B contains A which
is fixed, so B is tested against the error term.

The expected mean square for A is 
$$
Q(\alpha) + \frac{30}{15} \sigma_{\alpha\beta}^2 + \sigma^2
$$
and the expected mean square for A:B is 
$$
\frac{30}{15} \sigma_{\alpha\beta}^2 + \sigma^2.
$$
But as A is above A:B, the expected mean square for B is 
$$
\frac{30}{3} \sigma_{\beta}^{2} + \sigma^2
$$


### Fixed A,B

For the model where both A and B are fixed, again only the parentheses change in the diagram
```{r}
grViz(hasseDiagram(~ A*B,data=df))
```

In this case A, B and A:B are tested against the error term and the expected mean for A is 
$$
Q(\alpha) + \sigma^2
$$
the expected mean square for B is 
$$
Q(\beta) + \sigma^2
$$

and the expected mean square for A:B is
$$
Q(\alpha\beta) + \sigma^2
$$


## Nested ANOVA

Consider a simple nested design. Again suppose A has 5 levels, B has 3 levels and is nested in A,
and we take 2 replicate measurements.

### Random A,B

For the model where both A and B are random the diagram is

```{r}
grViz(hasseDiagram(~ A/B,random=c("A","B"),data=df))
```

Here B is nested within A so it is directly under A in the diagram, and

* A has $5$ levels, so it has superscript $5$ and its subscript is $5-1=4$.
* B has $3$ levels and is nested within A, so it is represented in the diagram as A:B and has
  superscript $5 \times 3 = 15$ with subscript $15-1-4=10$.
* The error term always has the total number of observations as its superscript, and its subscript
  is $30-1-4-10=15$.

In this case A is tested against B, and B is tested against the error term.

The expected mean square for A is
$$
\frac{30}{5} \sigma_{\alpha}^{2} + \frac{30}{15} \sigma_{\beta}^2 + \sigma^2
$$

and the expected mean square for B is
$$
\frac{30}{15} \sigma_{\beta}^2 + \sigma^2
$$

### Fixed A, Random B

When A is fixed

```{r}
grViz(hasseDiagram(~ A/B,random=c("B"),data=df))
```

Again A is tested against B, and B is tested against the error term.

The expected mean square for A is
$$
Q(\alpha) + \frac{30}{15} \sigma_{\beta}^2 + \sigma^2
$$

and the expected mean square for B is
$$
\frac{30}{15} \sigma_{\beta}^2 + \sigma^2
$$


## Complex Design

Consider a factorial design with two crossed factors A and B, with a third factor C nested within
combinations of A and B.  Suppose A has 5 levels, B has 3 levels and C has 4, and we take 2
replicate measurements.

```{r}
df <- expand.grid(A=factor(1:5),B=factor(1:3),C=factor(1:4),Rep=factor(1:2))
```

### Fixed A, B

If A and B are fixed and C is random

```{r}
grViz(hasseDiagram(~ (A*B)/C,random=c("C"),data=df))
```


Here

* A has $5$ levels, so it has superscript $5$ and its subscript is $5-1=4$.
* B has $3$ levels, so it has superscript $3$ and its subscript is $3-1=2$.
* A:B is constructed from A and B so its superscript is the product of the superscripts of A and B,
  $5 \times 3 = 15$, and its subscript is $15-1-4-2=8$.
* C is nested in A:B, so it appears in the diagram as A:B:C, its superscript is the the product of
  the superscripts of A and B and the levels in C $5 \times 3 \times 4 = 60$ and its subscript is
  $60-1-4-2-8=45$.
* The error term always has the total number of observations as its superscript, and its subscript
  is $120-1-4-2-8-45=60$.

In this case A, B and A:B are tested against C.

The expected mean square for A is
$$
Q(\alpha) + \frac{120}{60} \sigma_{\gamma}^{2}+\sigma^2
$$
The expected mean square for B is
$$
Q(\beta) + \frac{120}{60} \sigma_{\gamma}^{2}+\sigma^2
$$
The expected mean square for AB is
$$
Q(\alpha\beta) + \frac{120}{60} \sigma_{\gamma}^{2}+\sigma^2
$$
The expected mean square for C is
$$
\frac{120}{60} \sigma_{\gamma}^{2}+\sigma^2
$$

### Random A, B

If A, B and C are random
```{r}
grViz(hasseDiagram(~ (A*B)/C,random=c("A","B","C"),data=df))
```

In this case, A and B are tested against A:B, A:B is tested against C and C is tested against the
error term.


The expected mean square for A is
$$
\frac{120}{5} \sigma_{\alpha}^2 + \frac{120}{15} \sigma_{\alpha\beta}^2 + \frac{120}{60} \sigma_{\gamma}^{2}+\sigma^2
$$
The expected mean square for B is
$$
\frac{120}{3} \sigma_{\beta}^2 + \frac{120}{15} \sigma_{\alpha\beta}^2 + \frac{120}{60} \sigma_{\gamma}^{2}+\sigma^2
$$
The expected mean square for A:B is
$$
\frac{120}{15} \sigma_{\alpha\beta}^2 + \frac{120}{60} \sigma_{\gamma}^{2}+\sigma^2
$$
The expected mean square for C is
$$
\frac{120}{60} \sigma_{\gamma}^{2}+\sigma^2
$$


### Fixed A, Random B

If A is fixed and B and C are random
```{r}
grViz(hasseDiagram(~ (A*B)/C,random=c("B","C"),data=df))
```


Then A is tested against A:B, but A:B contains A so B is tested against C, and A:B is tested against 
C and C is tested against the error term.

The expected mean square for A is
$$
Q(\alpha) + \frac{120}{15} \sigma_{\alpha\beta}^2 + \frac{120}{60} \sigma_{\gamma}^{2}+\sigma^2
$$
The expected mean square for B is
$$
\frac{120}{3} \sigma_{\beta}^2 + \frac{120}{60} \sigma_{\gamma}^{2}+\sigma^2
$$
The expected mean square for A:B is
$$
\frac{120}{15} \sigma_{\alpha\beta}^2 + \frac{120}{60} \sigma_{\gamma}^{2}+\sigma^2
$$

The expected mean square for C is
$$
\frac{120}{60} \sigma_{\gamma}^{2}+\sigma^2
$$



## Quasi-F Ratios

Quasi-F ratios occur when there is no unique term to test against.  Consider a three way factorial
design where A, B and C are random, again with 5, 3 and 4 levels respectively and two replicate
measurements
```{r}
grViz(hasseDiagram(~ A*B*C,random=c("A","B","C"),data=df))
```

In this case there is no unique "highest" parenthesized term below A, so to test A we would need to
construct a quasi-F ratio.

The expected mean square for A is
$$
\frac{120}{5} \sigma_{\alpha}^2 + \frac{120}{15} \sigma_{\alpha\beta}^2 + 
\frac{120}{20} \sigma_{\alpha\gamma}^{2}+\frac{120}{60} \sigma_{\alpha\beta\gamma}^{2}+\sigma^2
$$
while the expected mean square for A:B is
$$
\frac{120}{15} \sigma_{\alpha\beta}^2 + \frac{120}{60} \sigma_{\alpha\beta\gamma}^{2}+\sigma^2
$$
the expected mean square for A:C is
$$
\frac{120}{20} \sigma_{\alpha\gamma}^2 + \frac{120}{60} \sigma_{\alpha\beta\gamma}^{2}+\sigma^2
$$
and the expected mean square for A:B:C is
$$
\frac{120}{60} \sigma_{\alpha\beta\gamma}^{2}+\sigma^2
$$
In this case to test A we can test against
$$
\operatorname{EMS}(\text{A:B})+\operatorname{EMS}(\text{B:C})-\operatorname{EMS}(\text{A:B:C})
=\frac{120}{15} \sigma_{\alpha\beta}^2 + 
\frac{120}{20} \sigma_{\alpha\gamma}^{2}+\frac{120}{60} \sigma_{\alpha\beta\gamma}^{2}+\sigma^2
$$


## References

Oehlert, G. W. (2010). A first course in design and analysis of experiments.
