---
title: "Newman"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Newman}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

## Block Interactions

Newman et al. (1997) discuss the inclusion of interactions with blocking factors in mixed model 
anova.  Here we reproduce the blue crab example from Newman et al. (1997) with the Hasse package.

```{r setup}
library(Hasse)
library(DiagrammeR)
```


## Data

The data from the blue crab predation example given by Newman et al. (1997) do not appear to be 
available and so we simulate data with an identical sums of squares decomposition 
```{r}
## Simulate a response that yields the given sums of squares
simulateSS <- function(formula,data,SS) {
  tms <- terms(formula)
  X <- model.matrix(delete.response(tms),data=data)
  Q <- qr.Q(qr(X))
  ks <- attr(X,"assign")
  err <- rnorm(nrow(Q))
  err <- err - Q%*%t(Q)%*%err
  err <- err/sqrt(sum(err^2))
  Q <- cbind(Q[,ks>0],err)
  ks <- c(ks[ks>0],max(ks)+1)
  bs <- rnorm(ncol(Q))
  bs <- bs*sqrt(SS/tapply(bs^2,ks,sum))[ks]
  data[,as.character(tms[[2]])] <- Q%*%bs
  data
}
```

Using the sums of squares reported for model 1 in Table 1 on page 1314, 
```{r}
## Simulate similar data
df <- expand.grid(
  Block=factor(c("1","3","5","7","9")),
  Habitat=factor(c("H1","H2")),
  Tether=factor(c("T1","T2")),
  Rep=1:10)
df <- simulateSS(Rate~Block*Habitat*Tether,data=df,
                 SS=c(11.473, 0.751, 0.36, 0.301, 0.409, 0.004, 0.137, 8.458))
head(df)
```

The simulated data reproduces the reported sums of squares for model 1 
```{r}
## Check sums of squares
fit <- aov(Rate~Block*Habitat*Tether,data=df)
summary(fit)
```
and model 2
```{r}
fit <- aov(Rate~Block+Habitat*Tether,data=df)
summary(fit)
```




## Block Interactions

First consider the case where interactions with the blocks are included in the model and construct 
a Hasse diagram treating the blocks as random
```{r}
## Hasse diagram when blocks interact with treatments
grViz(hasseDiagram(Rate~Block*Habitat*Tether,random="Block",data=df))
```

The Hasse diagram shows that 

* the habitat effect should be tested against the block by habitat interaction,
* the tether effect should be tested against the block by tether interaction, and
* the habitat by tether interaction should be tested against the block by habitat by tether 
  interaction

But the tests for the block effect and the block by habitat and block by tether interactions depend 
upon whether the restricted or unrestricted model is used.  For the restricted model
```{r}
## Correct tests from EMS (restricted model)
fit.ems <- aov(Rate~Block*Habitat*Tether,data=df)
mixed(fit.ems,random = "Block",restricted=T)
```

For the unrestricted model there is no test for the block effect (a quasi-F test is required) and 
all second order interactions are tested against the third order interaction.
```{r}
## Correct tests from EMS (unrestricted model)
fit.ems <- aov(Rate~Block*Habitat*Tether,data=df)
mixed(fit.ems,random = "Block",restricted=F)
```

The equivalent results for `aov` are obtained by placing all the random terms in the `Error` 
specification
```{r}
## Multistratum model
fit <- aov(Rate~Habitat*Tether+Error(Block+Block:Habitat+Block:Tether+Block:Habitat:Tether),
           data=df)
summary(fit)
```



### Additive Blocks

Alternately, consider the case where interactions where the block effects are assumed purely 
additive and construct a Hasse diagram
```{r}
## Hasse Diagram when blocks additive
grViz(hasseDiagram(Rate~Block+Habitat*Tether,random="Block",data=df))
```

In this case the diagram shows that all the effects should be tested against the error sums of 
squares 
```{r}
## Correct tests from EMS
fit.ems <- aov(Rate~Block+Habitat*Tether,data=df)
mixed(fit.ems,random = "Block")
```

The equivalent analysis with `aov` is given by
```{r}
## Multistratum model
fit <- aov(Rate~Habitat*Tether+Error(Block),data=df)
summary(fit)
```


## References

Newman, J. A., Bergelson, J., & Grafen, A. (1997). Blocking factors and hypothesis tests in ecology: 
is your statistics text wrong?. *Ecology*, 78(5), 1312-1320.

Oehlert, G. W. (2010). A first course in design and analysis of experiments.

